## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  print(a)
}else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
  print("small")
} else if (a > 5 & a <= 10) {
  print("medium")
} else if (a > 10 | a == 7) {
  print("large")
} else {
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}

a <- 5
if (a/2 != 0) {
  if (a == 5) {
    print("a is 5")
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  #Paste is a concatenation function
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

#change the first, third, and sixth elements to the value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)
mat

for (row in 1:2){
  for (col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10) {
  b <- b + 2
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
# start with a vector of temperatures
test_temp <- c(16.55, 26.94, 32.59, 8.58, 28.00, 28.54, 20.98, 21.17, 21.05, 18.77)

#Then we want a function that converts celcius to farenheit

c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) + 32
  return(temp_fahr)
}

c_to_f(test_temp)
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow = 2)
test_mat
#1 is for rows, 2 is for columns
apply(test_mat, 1, sum)
apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.55, 26.94, 32.59, 8.58, 28.00, 28.54, 20.98, 21.17, 21.05, 18.77), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
apply(test_temp_mat, 2, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1) Implement a function which convert the given boolean value into its string representation.

     boolean_to_string <- function(b){
       if (b == TRUE) {
         return("TRUE")
       } else {
         return("FALSE")
       }
     }
    ```

    ```{r}
    #2) The cockroach is one of the fastest insects. Write a function which takes its speed in km per hour and returns it in cm per second, rounded down to the integer (= floored).

    cockroach_speed <- function(s){
     cm <- as.integer(s*27.777778)
      return(cm)
    }
    ```

    ```{r}
    #3) It's pretty straightforward. Your goal is to create a function that removes the first and last characters of a string. You're given one parameter, the original string. You don't have to worry with strings with less than two characters.

    removeChar <- function(str){
      deleted_string <- substr(str, 2, nchar(str)-1)
      return(deleted_string)
    }
    ```

    ```{r}
    #4) Simple, remove the spaces from the string, then return the resultant string.

    no_space <- function(x){
      new_str <- gsub(" ", "", x)
      return(new_str)
    }
    ```

    ```{r}
    #5) This kata is about multiplying a given number by eight if it is an even number and by nine otherwise.

    simple_multiplication <- function(number) {
      if(number %% 2 == 0) {number * 8} else {number * 9}
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}

    bzbz <- function(num){
      if (num %% 3 == 0 & num %% 5 == 0) {
        return("fizzbuzz")
      } else if (num %% 3 == 0) {
        return("fizz")
      } else if (num %% 5 == 0) {
        return("buzz")
      } else {
        return("baiiiii")
      }
    }

    bzbz(15)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    my_stair <- "X"
    mario <- function(stair){
      for (i in 1:stair){
      print(strrep(my_stair, i))
    }
    }
    mario(4)
    ```
